On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-24 paper/application.tex
finished revising 6.2

\begin{itemize}
  \item $\dot{M}$, the global radial mass accretion rate;
  \item $Q_0$, the value of the 2D Toomre parameter where the disk is
    gravito-turbulent;
  \item $T_\mathrm{irr}$, the irradiation or floor temperature;
  \item $\alpha_m$, the dimensionless viscosity associated with other
    sources of turbulence.
\end{itemize} 
These are used in the statements of thermal equilibrium, mass and
angular momentum conservation, together with an opacity law, 
to construct a global disk model with surface density $\Sigma(R)$ and
temperature $T(R)$ where $R$ is the global cylindrical radius from the
star. \citepalias[See][for details.]{rafikov15} This gives the $Q(R)$
value required for input into our linear framework. We derive other
dimensionless parameters below.

Note that we are considering the local stability at each radius. We
are thus neglecting the slow, background viscous radial flow in the
linear problem.

%proper treatment of cooling  in 3d
%vertically isothermal basic state for simplicity (also appropriate for
%irradiated disks) --- this formally requires zero viscosity  
%or have viscosity but assume it does not affect basic state (dominated
%by irradiation) but affect perturbations

\subsection{Effective $\alpha$}
In a steady, viscously accreting Keplerian disk we have,
approximately, $\nu\Sigma = \dot{M}/3\pi$. Hence
\begin{align}
  \alpha = \frac{\dot{M}}{3\pi}\frac{\Omega(R)}{c_{s}^2(T)\Sigma(R)},  
\end{align} 
which sets the viscosity coefficient to be used at each radius.

\subsection{PPD cooling}
Energy loss in \citetalias{rafikov15}'s model is given by 
\begin{align}\label{real_cool}
  \Lambda = \frac{2\sigma}{f(\tau)}\left(T^4 - T_\mathrm{irr}^4\right)
\end{align}
per unit area, 
\begin{align}
  f(\tau) = \tau + \frac{1}{\tau}, \label{ftau} 
\end{align}
and
\begin{align}
  \tau = \kappa_d(T)\Sigma
\end{align}
is the optical depth and recall Eq. \ref{opacity_law} 
is our opacity model. Eq. \ref{ftau} accounts for cooling in the
optically-thin ($\tau\ll 1$) and optically-thick ($\tau\gg1$)
regimes.

Since we are interested in the linear response of the system, the
corresponding cooling parameters $\beta$ and $\theta$ are found by
comparing the linearized form of the cooling function
(Eq. \ref{real_cool}) to that of the model function
(Eq. \ref{beta_cool}).  
%{\bf as opposed to comparing their unperturbed forms, as done in literature}
Linearizing Eq. \ref{real_cool} gives   
\begin{align}\label{linear_real_cool}
  (\gamma-1)\frac{\delta\Lambda}{\Sigma} = \frac{2\sigma(\gamma-1)
    T^4C_1}{f(\tau)c_{s}^2\Sigma}\left(\frac{\delta P}{\Sigma} -
  \frac{C_2}{C_1}c_{s}^2\frac{\delta\Sigma}{\Sigma}\right), 
\end{align}
where
\begin{align}
C_1(\tau, T) &= 4 - b\times g(\tau, T),\\ 
C_2(\tau, T) &= 4 + (1-b)\times g(\tau, T),\\ 
  g(\tau, T) &= \left( \frac{\tau^2-1}{\tau^2+1}\right)\left(1 -
  \frac{T_\mathrm{irr}^4}{T^4}\right). 
\end{align}
Comparing Eq. \ref{linear_real_cool} with Eq. \ref{beta_cool}, we identify
\begin{align}
  &\beta = \frac{f(\tau)c_s^2\Sigma\Omega}{2\sigma(\gamma-1)C_1T^4},\\
  &\theta = \frac{C_2}{C_1}, 
\end{align}
to be used in the 2D linear framework.

This `realistic' beta cooling prescription can behave differently to 
our previous, simple beta cooling (Eq. \ref{beta_cool}) for the same
physical system. In particular, for a non-irradiated disk
($T_\mathrm{irr}=0$), simple beta 
cooling has  $\theta=0$, but realistic beta cooling has $\theta\neq
0$. However, fully-irradiated disks with $T=T_\mathrm{irr}$ can be
modeled by simple beta cooling since in both cases $\theta=1$.

>>>>

\begin{itemize}
  \item $\dot{M}$, the global radial mass accretion rate;
  \item $Q_0$, the value of the 2D Toomre parameter where the disk is
    gravito-turbulent;
  \item $T_\mathrm{irr}$, the irradiation or floor temperature;
  \item $\alpha_m$, the dimensionless viscosity associated with other
    sources of turbulence, such as magneto-rotational instabilities
    (MRI), see also \S\ref{visc_gi} and \S\ref{MHD}. 
\end{itemize} 
These are used in the statements of thermal equilibrium, mass and
angular momentum conservation, together with an opacity law, 
to construct a global disk model with surface density $\Sigma(R)$ and
temperature $T(R)$ where $R$ is the global cylindrical radius from the
star. \citepalias[See][for details.]{rafikov15} This gives the $Q(R)$
value required for input into our linear framework. We derive other
dimensionless parameters below.

Note that we are considering the local stability at each radius. We
are thus neglecting the slow, background viscous radial flow; as well
as any global gravitational instabilities \citep{lodato05}. {\bf more refs?}

\subsection{Effective $\alpha$}
In a steady, viscously accreting Keplerian disk 
with constant $\dot{M}$ we have,
approximately, $\nu\Sigma = \dot{M}/3\pi$. Hence
\begin{align}
  \alpha = \frac{\dot{M}}{3\pi}\frac{\Omega(R)}{c_{s}^2(T)\Sigma(R)},  
\end{align} 
which sets the viscosity coefficient to be used at each radius.

\subsection{PPD beta cooling}
Energy loss in \citetalias{rafikov15} is given by 
\begin{align}\label{real_cool}
  \Lambda = \frac{2\sigma}{f(\tau)}\left(T^4 - T_\mathrm{irr}^4\right)
\end{align}
per unit area, 
\begin{align}
  f(\tau) = \tau + \frac{1}{\tau}, \label{ftau} 
\end{align}
and
\begin{align}
  \tau = \kappa_d(T)\Sigma
\end{align}
is the optical depth and recall Eq. \ref{opacity_law} 
is our opacity model where $\kappa_d\propto T^b$ with $b=2$. 
Eq. \ref{ftau} accounts for cooling in the optically-thin ($\tau\ll
1$) and optically-thick ($\tau\gg1$) regimes.

Note that Eq. \ref{real_cool} is a beta cooling prescription because
it is an explicit function of the thermodynamic states. However, we
formulated the linear problem with beta cooling of the functional form
given by Eq. \ref{beta_cool}. In order to adapt the developed
framework to the above PPD beta cooling function, we
need to identify the $\beta$ and $\theta$ parameters that enter the 
linearized equations (Eq. \ref{linear_beta}).

%Since we are interested in the linear response of the system, the
%corresponding cooling parameters $\beta$ and $\theta$ are found by
%comparing the linearized form of the cooling function
%(Eq. \ref{real_cool}) to that of the model function
%(Eq. \ref{beta_cool}).

%{\bf as opposed to comparing their unperturbed forms, as done in literature}
Linearizing Eq. \ref{real_cool} gives   
\begin{align}\label{linear_real_cool}
  (\gamma-1)\frac{\delta\Lambda}{\Sigma} = \frac{2\sigma(\gamma-1)
    T^4C_1}{f(\tau)c_{s}^2\Sigma}\left(\frac{\delta P}{\Sigma} -
  \frac{C_2}{C_1}c_{s}^2\frac{\delta\Sigma}{\Sigma}\right), 
\end{align}
where
\begin{align}
C_1(\tau, T) &= 4 - b\times g(\tau, T),\\ 
C_2(\tau, T) &= 4 + (1-b)\times g(\tau, T),\\ 
  g(\tau, T) &= \left( \frac{\tau^2-1}{\tau^2+1}\right)\left(1 -
  \frac{T_\mathrm{irr}^4}{T^4}\right). \label{g_def}
\end{align}
Comparing Eq. \ref{linear_real_cool} with the linearized form of the
standard cooling function, Eq. \ref{linear_beta}, we identify
\begin{align}
  &\beta = \frac{f(\tau)c_s^2\Sigma\Omega}{2\sigma(\gamma-1)C_1T^4},\label{real_beta}\\
  &\theta = \frac{C_2}{C_1},\label{real_theta} 
\end{align}
to be used in the 2D dispersion relation (Eq. \ref{thindisk}).

Eq. \ref{real_beta} still represents a physical (dimensionless) cooling
time, and is consistent with previous definitions of realistic cooling
times within factors of order unity \citep[e.g.][their
Eq. 2]{kratter10}.

However, here $\theta$ is an effective or generalized
irradiation level. 
%no longer a direct measure of
%irradiation. 
By Eq. \ref{real_theta}, $\theta$ is now related to the true
irradation temperature $\tirr$ through the function $g$ given by 
Eq. \ref{g_def}. This is
important because a physically non-irradiated disk with $\tirr=0$ will
still have a non-zero effective irradiation, $\theta\neq0$. 
%On the
%other hand, fully irradiated disks with  $T=T_\mathrm{irr}$ have 
%$\theta=1$, as is in the previous problem formulation
%(Eq. \ref{tirr_def}).

%the gammie-type cooling function does not represent realistic
%cooling. because real cooling depends on sigma and pressure but gammie
%cooling only depends on pressure

### 2016-01-24 paper/model3d.tex
finished revising 6.2

\subsubsection{Beta cooling}
In this case the energy loss per unit volume is specified 
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}. 
\end{align}
Recall from Eq. \ref{beta_def} that 
%\begin{align}
  $\tcool = \beta\Omega^{-1}$ is 
%\end{align}
the cooling timescale with $\beta$ a constant input parameter. Here, 
\begin{align}
  \tirr = \theta T_\mathrm{eq}
\end{align}
is a reference temperature field parameterized as a constant 
fraction $\theta$ of the equilibrium temperature
$T_\mathrm{eq}$. Physically, $\tirr$ is the floor temperature set by,
for example, stellar irradiation. We will thus refer to $\theta$ as
the `irradiation level'. %stellar illum or env rad

>>>>

\subsubsection{Beta cooling}\label{beta_cool_model}
In this case the energy loss per unit volume is specified 
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}. 
\end{align}
Recall from Eq. \ref{beta_def} that 
%\begin{align}
  $\tcool = \beta\Omega^{-1}$ is 
%\end{align}
the cooling timescale with $\beta$ a constant input parameter. Here, 
\begin{align}\label{tirr_def}
  \tirr = \theta T_\mathrm{eq}
\end{align}
is a reference temperature field parameterized as a constant 
fraction $\theta$ of the equilibrium temperature
$T_\mathrm{eq}$. Physically, $\tirr$ is the floor temperature set by,
for example, stellar irradiation. We will thus refer to $\theta$ as
the `irradiation level'. %stellar illum or env rad

### 2016-01-24 paper/paper3d.tex
halfway through version 2 of 2d results
finished revising 6.2

Despite this, the Toomre parameter and its associated
dispersion relation are still widely applied to viscous, cooling
accretion disk models \citep[e.g.][]{cossins09, kimura12}, {\bf more refs?}
even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
derived from inviscid disks without cooling.

@@

We will quantify how cooling removes pressure support to enable GI
even when $Q>1$. For example, if perturbations are
cooled towards zero tempertaure, we find the above dispersion relation is
modified to read 
\begin{align}\label{cool_dr}
s^2 = 2\pi G\Sigma |k| - \kappa^2 - \gamma \left(\frac{\tcool s}{1 +
  \tcool s}\right)c_s^2k^2,
\end{align}
where $\gamma$ is the adiabatic index. Cooling increases the growth
rate by reducing the magnitude of the 
pressure term. In fact, for $t_c\in(0,\infty)$ the formal condition
for instability is  
\begin{align*}
  2\pi G \Sigma |k| > \kappa^2,
\end{align*}
which is what would be obtained from Eq. \ref{classic_dr} with 
pressure neglected ($c_s\to 0$). This condition does not actually
depend on the cooling rate, and instability is
possible for any finite $Q$. %at sufficiently small scales 
This shows that the mere presence of
cooling changes the qualitative nature of GI compared to the simple
Toomre condition.   
%{\bf numerical implications? if using standard beta cooling, then as
%  long as it's there then sufficiently small scales will be
 % unstbale. can access those scales with high resolution. defer to
 % discussion section}

@@

For an isothermal, viscous, self-gravitating disk in 2D,
\cite{gammie96} finds the approximate disperion relation   
\begin{align}
s \simeq \frac{\nu k^2\left(2\pi G\Sigma |k| -
  c_s^2k^2\right)}{\kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|}. 
\end{align}
Assuming $Q>1$, instability occurs if 
\begin{align*}
  2\pi G \Sigma > c_s^2|k|,
\end{align*}
provided that $\nu k^2\neq0$. This condition for viscous GI is 
identical that would be obtained from Eq. \ref{classic_dr} with
rotation neglected ($\Omega\to 0$). That is, a classically stable
disk can be made unstable by viscosity as it effectively removes
rotational stabilization \citep{lynden-bell74}.

>>>>

Despite this, the Toomre parameter and its associated
dispersion relation are still widely applied to viscous, cooling
accretion disk models \citep[e.g.][]{cossins09, kimura12}, {\bf more refs?}
even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
derived from inviscid disks without cooling. 
%e.g. using classic DR to determine what scales need to be resolved

@@

We will quantify how cooling removes pressure support to enable GI
even when $Q>1$. For example, if perturbations are
cooled towards zero tempertaure, we find the above dispersion relation is
modified to read 
\begin{align}\label{cool_dr}
s^2 = 2\pi G\Sigma |k| - \kappa^2 - \gamma \left(\frac{\tcool s}{1 +
  \tcool s}\right)c_s^2k^2,
\end{align}
where $\gamma$ is the adiabatic index. Cooling increases the growth
rate by reducing the magnitude of the 
pressure term. In fact, for $t_c\in[0,\infty)$ the formal condition
for instability is  
\begin{align*}
  2\pi G \Sigma |k| > \kappa^2,
\end{align*}
which is what would be obtained from Eq. \ref{classic_dr} with 
pressure neglected ($c_s\to 0$). This condition does not actually
depend on the cooling rate, and instability is
possible for any finite $Q$. %at sufficiently small scales 
This shows that the mere presence of
cooling changes the qualitative nature of GI compared to the simple
Toomre condition.   
%{\bf numerical implications? if using standard beta cooling, then as
%  long as it's there then sufficiently small scales will be
 % unstbale. can access those scales with high resolution. defer to
 % discussion section}

@@

For an isothermal, viscous, self-gravitating disk in 2D,
\cite{gammie96} finds the approximate disperion relation   
\begin{align}
s \simeq \frac{\nu k^2\left(2\pi G\Sigma |k| -
  c_s^2k^2\right)}{\kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|}. 
\end{align}
Assuming $Q>1$, instability occurs if 
\begin{align*}
  2\pi G \Sigma > c_s^2|k|,
\end{align*}
provided that $\nu k^2\neq0$. This condition for viscous GI is 
identical that would be obtained from Eq. \ref{classic_dr} with
rotation neglected ($\Omega\to 0$). That is, a classically stable
disk can be destabilized by viscosity as it effectively removes
rotational stabilization \citep{lynden-bell74}.
